---
title: "Estimation_plots_with_R"
author: "Thomas Solomon"
date: "2025-02-21"
output: html_document
---

# Data Analysis with Bootstrap-coupled ESTimation (DABEST)

<p> Data Analysis with Bootstrap-coupled ESTimation (DABEST) was developed by Ho <em>et al</em>. (2019) <em>Nature Methods</em> (<a target="_blank" href="https://www.biorxiv.org/content/10.1101/377978v2.full">https://www.biorxiv.org/content/10.1101/377978v2.full</a> see <a target="_blank" href="https://rdcu.be/bHhJ4">full-text here</a>).</p>

<p> DABEST creates an <em>estimation plot</em>, aka a "Gardner–Altman plot", which shows the distribution of mean differences (or effect sizes) between conditions. A "Gardner–Altman plot" presents all data points as a swarm plot and the mean difference(s), or effect size(s), as a bootstrap 95% confidence interval.</p>

<p> See the manual for using DABEST in R at <a target="_blank" href="https://acclab.github.io/dabestr/">https://acclab.github.io/dabestr/</a>.</p>

<p> See the original R code from Ho <em>et al</em>. (2019) on <a target="_blank" href="https://github.com/ACCLAB/dabestr">GitHub</a>.</p>

# Install packages
``` {r}
# Install  packages 
library(tidyverse)
install.packages('dabestr')
library(dabestr)
```
# DABEST using simulated data
## Simulate PAIRED data
``` {r}
# Simulate some data in R
set.seed(123)
n <- 30
C <- rnorm(n, mean = 3.0, sd = 0.5)  # Control group
T <- rnorm(n, mean = 3.3, sd = 0.5)  # Test group

# Create an "id" column for paired observations. Remove if data are UNPAIRED.
ids <- 1:n

df_simulate <- data.frame(
  id = rep(ids, 2),               # each id appears twice, one for each group
  group = rep(c("Control", "Test"), each = n),
  value = c(C, T)
)

# Examine the dataframe structure
str(df_simulate)
```
``` {r}
# Summarise the dataframe 
summary(df_simulate)
```
``` {r}
# View the dataframe head
head(df_simulate)
```
``` {r}
# View the dataframe tail
tail(df_simulate)
```
``` {r}
# Create histograms to visualize the distribution of data in each group
ggplot(df_simulate, aes(x = value, fill = group)) +
  geom_histogram(alpha = 0.5, position = "identity") +
  labs(title = "Distribution of Values by Group", x = "Value", y = "Frequency") +
  facet_wrap(~group)
```
## Load the data into dabestr
``` {r}
# Load the data into `dabest` with 5000 bootstrap samples. Use paired="baseline" to indicate that the data are paired and id_col=id to help DABEST find the paired identiers
# NOTE: By default, 5000 resamples will be used to generate the effect size bootstraps
# NOTE: When only 2 paired data groups are involved, assigning either “baseline” or “sequential” to paired will give you the same numerical results.
dabest_simulate <- load(df_simulate, x = group, y = value, idx = c("Control", "Test"), paired = "baseline", id_col = id, resamples = 5000)
```
## Mean difference for PAIRED data
``` {r}
# calculate the mean difference (use median＿diff for the median difference)
dabest_simulate.mean_diff <- mean_diff(dabest_simulate)
print(dabest_simulate.mean_diff)
```
### Gardner-Altman estimation plot
``` {r}
# build the estimation plot (Gardner-Altman plot) for the mean difference
dabest_plot(dabest_simulate.mean_diff, raw_marker_size = 0.5, raw_marker_alpha = 0.3)
```
### Cumming estimation plot
``` {r}
# Instead of a Gardner-Altman plot, you can generate a Cumming estimation plot by setting float_contrast=False in the plot() method.
dabest_plot(dabest_simulate.mean_diff, float_contrast = FALSE, raw_marker_size = 0.5, raw_marker_alpha = 0.3, contrast_ylim = c(-0.3, 1.3))
```
